\section{Background and Motivation}
\label{sec:background}

\subsection{Real-world agent work}

We use \emph{real-world agent work} to mean work in which agents help users
manage concrete tasks rather than only produce isolated model outputs. Such
work may involve local state, external repositories, remote terminals, GUI and
TUI controls, model calls, package releases, documents, and human decisions.
The product problem is not only how to run an agent. It is how to keep the
state of the work legible over time.

Long-lived work introduces a persistence problem. Users and agents need to ask:
what happened, what source observed it, what evidence supports it, what is
missing, and what action is safe now? A timeline view is a natural interface
for these questions, but only if the view makes its own assumptions inspectable.

\subsection{The global-clock temptation}

Distributed systems theory has long shown that time and ordering are not
trivial in distributed computation. Lamport's happened-before relation
separates causal order from physical clock order \cite{lamport1978time};
vector-clock style systems make concurrency explicit \cite{mattern1989virtual};
and distributed snapshots show that a useful global state is a constructed
consistent cut rather than a simultaneous physical observation
\cite{chandy1985snapshots}.
Yet many product surfaces still sort mixed-source facts primarily by wall-clock
timestamps and present the result as a single history. That is often sufficient
for casual logs, but it is not sufficient for recovery, audit, or cooperation
among capable participants.

The problem is sharper in agent systems because the user may not have the time
or expertise to manually diagnose missing evidence. A local product must help
both the human and the agent understand what the view means.

\subsection{The immortal-observer temptation}

A system can reject a universal clock while still assuming that its observer is
stable. That assumption also fails in long-lived work. Processes restart,
controllers are replaced, frontends and runtimes upgrade independently, and
physical execution may survive the component that displayed it. A process
identifier or heartbeat shows placement and liveness; it does not prove that a
replacement accepted the same authority, reconstructed the same projection, or
understands the same policy version.

Failure-detector theory similarly separates suspicions about process failure
from stronger system guarantees \cite{chandra1996failure}. In the product
setting, detecting that a process disappeared is only the start of recovery;
it does not prove what authority its successor may safely claim.

This creates a distinction between a stable logical observer and one physical
or software \emph{incarnation} of that observer. A replacement may legitimately
continue the logical perspective, but only after it establishes an accepted
cut, rejects stale authority, rebuilds derived state, and records the semantic
version under which it will project the facts. Otherwise the product should
show a discontinuity or degraded view rather than imply seamless recovery.

\subsection{KFD foundation}

The KFD foundation used by Kungfu can be summarized as:

\begin{quote}
KFD-1: facts must not drift.\\
KFD-2: trust must start from facts.\\
KFD-3: cooperation must start from trusted value.
\end{quote}

KFD-4 applies this foundation to timelines:

\begin{quote}
KFD-4: timelines must declare their observer.
\end{quote}

Under this framing, a timeline view is trustworthy only when it is grounded in
non-drifting facts, exposes enough evidence for trust assessment, and helps
participants cooperate through transparent value rather than hidden pressure or
opaque authority \cite{kfd2026observer}.

\subsection{Why a causal object is needed}

A projection rule alone does not define the unit that can be accepted, moved,
verified, or rejected. Raw journal pages are physical append blocks; sessions
and runs are product vocabulary; individual frames are too small to provide a
portable closure boundary. Kungfu therefore introduces an \emph{Episode}: a
bounded causal segment containing the manifest, frame references, content
commitments, provenance, schemas, dependencies, and verification evidence
needed to inspect one unit of work \cite{kungfu2026episode}.

Episodes do not make the timeline absolute. They make its inputs explicit. A
view can name which Episodes it accepted, preserve their internal and declared
cross-Episode causality, and then apply observer policy only where the causal
facts leave an ordering choice.

Episodes also survive observer replacement. This matters because fact
continuity should not depend on the lifetime of the process that currently
projects or presents those facts. Observer continuity is then a separately
verifiable claim over stable authority rather than a second owner of history.
