\section{Discussion and Limitations}
\label{sec:discussion}

\subsection{Not relativism}

Declaring an observer does not mean that any story is acceptable. Facts,
causality, hashes, manifests, schemas, cuts, and verification results remain
load-bearing. The rule rejects an undeclared view from nowhere, not the
existence of facts. Two observers may order concurrent facts differently; they
may not silently invert known causality or invent evidence.

\subsection{The observer is neither omniscient nor immortal}

The first version of the proposal emphasized that no observer has an absolute
global vantage point. Runtime implementation exposed a second requirement: an
observer also has a history. A logical identity can outlive a process, but the
successor must not inherit authority or semantics by name alone.

This refinement is useful beyond Kungfu. Long-lived local applications routinely
restart coordinators, reconnect clients, replace caches, and upgrade binaries.
Without incarnation metadata, a product may preserve visual continuity while
silently changing its accepted cut or interpretation. The continuity witness
makes that transition inspectable. It does not require one specific runtime
topology; embedded, process-hosted, and remote implementations can provide
different evidence for the same surface obligation.

\subsection{Continuity is layered}

Fact-authority continuity, observer continuity, and presentation continuity
fail independently. Treating one as proof of the others creates common but
dangerous shortcuts:

\begin{itemize}[leftmargin=*]
  \item a live PID does not prove that the process owns current authority;
  \item a reattached terminal does not prove task completion or evidence quality;
  \item an intact ledger does not prove that the current projection reached it;
  \item a successful software download does not prove that a new runtime may
        safely activate; and
  \item a familiar GUI does not prove unchanged policy semantics.
\end{itemize}

Making these claims separate may expose more recovering or action-required
states. That is a product cost, but hiding them would exchange usability today
for unauditable recovery later. Dual-first interfaces can reduce the burden by
giving agents the same structured state that humans see in the product.

\subsection{Not universal consensus}

Some systems require consensus, total ordering, or a central sequencer. The
observer-declared timeline rule does not forbid those designs. It says that when
a product presents a timeline to a human or agent, the view should still declare
the accepted facts, cut, projection policy, and observer incarnation behind it.
A consensus result can itself be one accepted fact source.

\subsection{Adopter responsibility}

A schema can standardize vocabulary, but it cannot prove an adopter's runtime
correctness by itself. An adopter that claims a concrete timeline is complete,
fresh, continuous, or causally valid must provide product-specific evidence.
This is why KFD-4 should be assessed through KFD-2-style trust claims when used
as a release or product gate.

\subsection{What the prototype establishes}

The current prototype establishes that the proposed view can stand on a
nontrivial fact and single-host continuity substrate: bounded causal objects,
append-only authority, safe-capability assessment, exact-cut readiness,
generation and epoch fencing, stale-authority rejection, explicit recovery
states, and versioned runtime selection. It also demonstrates that different
evidence revisions can remain honest instead of being retroactively merged into
one stronger result.

It does not establish that observer declaration by itself improves human
recovery time or agent decision quality. Nor does it establish a distributed
consistency protocol, conflict resolution between independently written
authority roots, cross-host continuity, or a total order acceptable to every
observer.

\subsection{Implementation and evidence limits}

The Episode numerical baseline is metadata-only, uses a single reported seed at
the largest scale, and does not cover realistic payload volume, deep dependency
DAGs, distributed writers, long soak, storage exhaustion, or every publication
crash point. The retained runtime-activation result is Darwin arm64 and
single-host. Peer continuity has a passing local macOS process campaign but not
retained Linux and Windows promotion evidence. The upgrade implementation has
contract and synthetic coverage but not complete signed-platform and native
fault-campaign evidence.

Most importantly, the first-class observer policy object, accepted range
synchronization, multi-source projection command, dedicated Episode projection,
exported observer declaration, continuity witness export, and projection-level
causal fsck have not yet landed. These are not incidental implementation
details. They are the remaining bridge between a qualified substrate and an
evaluated multi-machine observer timeline.
