\section{Conclusion}

The KFD Foundation Triad preserves a shared work world through a recursive
kernel: facts support bounded trust, trust supports inspectable cooperation,
and cooperation creates consequences that return as new facts. Operating that
kernel across multiple participants reveals a necessary boundary. Stable facts
do not produce a view from nowhere.

KFD-4 follows from that pressure. A perspective-bearing timeline must declare
the observer, accepted fact cut, projection policy, causal constraints,
consequence position, known gaps, and verification evidence that produced the
view. KFD-1 prevents the projection from silently changing the underlying
reference. KFD-2 requires the view's warrant to remain reconstructable. KFD-3
requires its value, constraints, and safe action boundary to be inspectable to
the participant expected to use it. Observer declaration is therefore not a
fourth slogan beside the triad. It is one mechanism required when the triad is
used by participants who cannot occupy every position at once.

The core rule also exposes a second-order problem: an observer implementation
may fail or change. A logical observer can outlive one process, but continuity
between incarnations must be established through accepted cuts, policy and
runtime identity, recovery evidence, and explicit degradation rather than a
reused name or familiar interface.

Kungfu provides bounded implementation evidence beneath these claims. Episodes
realize accepted causal units with append-only authority and safe-capability
evidence. Exact-cut runtime readiness, generation and coordinator-epoch
fencing, Peer rebootstrap, explicit recovery states, and immutable runtime
selection show how one observer incarnation can change without silently
acquiring stale authority. These mechanisms instantiate the model; they are
not required to understand it, and their qualification does not certify the
complete observer-level result.

That result remains open. It requires first-class observer declarations,
accepted source ranges, reproducible cross-source projection, exported
continuity evidence, projection-level causal verification, and experiments in
which humans and agents reconstruct the same declared view and use it to choose
or refuse a next action. Preserving this boundary is part of the argument:
trust in a timeline must start from the facts available, while cooperation must
start from a view whose perspective and consequences are visible.
