\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[protrusion=true,expansion=false]{microtype}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{array}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=black,
  urlcolor=black
}

\title{Observer-Declared Timelines for Real-World Agent Work}
\author{Keren Dong\\Kungfu Origin Technology Limited\\\href{mailto:keren.dong@kungfu.link}{keren.dong@kungfu.link}}
\date{Draft: July 2026}

\begin{document}

\maketitle

\begin{abstract}
Software systems that manage real-world agent work increasingly combine facts
from local machines, remote sessions, repositories, user interfaces, command
line tools, model providers, and human decisions. These systems often present a
single timeline even though the underlying facts come from multiple clocks,
locations, and participants. This paper argues that such timelines should not
be presented as views from nowhere. A useful timeline should declare its
logical observer, observer incarnation, accepted facts, projection policy,
causal constraints, continuity evidence, and degraded evidence state. We call
this design rule \emph{observer-declared timelines}.
The rule is motivated by KFD-4, a Kungfu procedure derived from the foundation
that facts must not drift, trust must start from facts, and cooperation must
start from trusted value. We define a timeline as a deterministic projection
over accepted \emph{Episodes}: bounded causal objects whose manifests, evidence,
dependencies, and verification state can be inspected independently. We extend
the model with observer incarnations and continuity witnesses: a restarted or
upgraded process does not inherit perspective merely because it has the same
name. We then report a Kungfu implementation path comprising append-only fact
authority, exact-cut projection readiness, generation- and epoch-fenced Peer
recovery, immutable runtime selection, explicit recovery states, and
source-bound qualification. A historical metadata baseline retained and
re-read 100,000 Episodes and 200,000 semantic records without a correctness
violation; newer single-host evidence exercises exact-cut activation and
process recovery. These results establish a fact and continuity substrate, not
an end-to-end validation of observer-relative multi-machine projection.
First-class observer policy, accepted-range sync, reproducible cross-source
ordering, and projection-level causal verification remain future work.
\end{abstract}

\input{sections/01-introduction}
\input{sections/02-background}
\input{sections/03-model}
\input{sections/04-design}
\input{sections/05-evaluation}
\input{sections/06-related-work}
\input{sections/07-discussion}
\input{sections/08-conclusion}

\bibliographystyle{plain}
\bibliography{references}

\end{document}
